On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-04-04 manuscripts/PNAS/chazthm_pnas.tex
SR up to mid pg 4

% Please add here a significance statement to explain the relevance of your work
%\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership. The Significance Statement appears in the paper itself and is required for all research papers.}

@@

% Please include corresponding author, author contribution and author declaration information
\authordeclaration{The authors declare no conflict of interest.}
%\equalauthors{\textsuperscript{1}C.G. proved Theorem 1 and C.H. proved Theorem 2.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: cjg@berkeley.edu, chillar@msri.org}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Sparse linear coding, dictionary learning, uniqueness, compressed sensing}

@@

% It has remained unknown, however, when the dictionary learning problem is well-posed (as per Hadamard \cite{Hadamard1902}) in general -- that is, independent of all methodological idiosyncracies.
%[**TODO** make sure the claim here is right...or would it be punchier to talk about the fact that everyone always gets Gabors on natural images?]
%- algos with proven recovery of generating dictionary
%- analyses of when target dictionary is local optimizer of natural recovery criteria
%- efficient algos that globally solve DR; detailed models differ but all assume each col has bounded sparsity and nonzero coefs have sub-Gaussian magnitudes

More formally, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our combined worst-case uncertainty in  measuring $\mathbf{A}\mathbf{x}_i$.
The precise mathematical problem addressed here is the following.
% We address the uniqueness properties for the following linear coding question in signal processing.

\begin{problem}[Sparse linear coding]\label{InverseProblem}
Find a matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^m$ such that $\|\mathbf{z}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \eta$ for all $i$.
\end{problem}

Note that any particular solution $(\mathbf{B}, \mathbf{\hat x}_1 \ldots, \mathbf{\hat x}_N)$ to this problem gives rise to an orbit of other solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal.  
We therefore consider whether solutions to Prob.~\ref{InverseProblem} are unique only up to this ambiguity.  More specifically, 
we seek conditions for when sparsely coded datasets have a stable unique representation.

% More specifically,  we seek sufficient conditions on when sparsely coded datasets have a stable unique representation.

% We therefore consider only whether solutions to Problem~\ref{InverseProblem} are unique up to this equivalence \cite{Li15}

% Note that by formulating the problem in this way we have introduced the following ambiguity. Any particular solution represents . Alternatively, one may seek a unique solution (in the usual sense) in the quotient space induced by the action of the transformation group of matrices $PD$.

%  noiseless case $\eta = 0$

%  We introduce the following terminology to handle $\eta > 0$. 
\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse linear representation in $m$ dimensions} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. %Equivalently, $\mathbf{Y} = \mathbf{A}\mathbf{X}$ for $\mathbf{Y}$ and $\mathbf{X}$ having the $\mathbf{y}_i$ and $\mathbf{x}_i$ as their columns, respectively. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon,\ \ \   \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}
% Kato mentions a distance between n-tuples that minimizes the maximum difference over all possible orderings of one of the n-tuples.
% Make just 'sparse representation in m-dimensions? (i.e. with k unspecified?)

To see how Def. \ref{maindef} directly relates to Prob. \ref{InverseProblem}, suppose that $Y$
% \mbox{$Y = \{\mathbf{A} \mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$} 
has a stable $k$-sparse linear representation and fix $\delta_1, \delta_2$ to be the desired recovery accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then from the triangle inequality, it follows that any matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N$ solving Prob.~\ref{InverseProblem} are necessarily close to the original dictionary $\mathbf{A}$ and codes $\mathbf{x}_i$.

%Our results, outlined in Sec.~\ref{Results}, include an explicit form for $\varepsilon(\delta_1, \delta_2)$ in Thm.~\ref{DeterministicUniquenessTheorem}.

% We ask here: \emph{When does $Y$ have a stable $k$-sparse representation?}

@@

Our main technical finding is that a sufficient quantity of measurements generated from any dictionary satisfying the spark condition 
has a stable sparse representation.  More specifically, such dictionaries are uniquely identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $(n, m, k)$ satisfy CS inequality \eqref{CScondition}, in almost all cases the dictionary learning problem is well-posed given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies the spark condition. Our results  also apply when only an upper bound on the number of dictionary elements is known. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of sparse dictionary learning.  % (Thm.~\ref{DeterministicUniquenessTheorem2})

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including our main tool from combinatorial matrix analysis (Lem.~\ref{MainLemma}). All other proofs are relegated to the Supplemental. 
%Our approach is a refinement of the one found in \cite{Hillar15} to handle noise and to reduce the sufficient number of samples.
%for all $k < m$ from $N=k{m \choose k}^2$ to $N = m + m(k-1){m \choose k}$. 
Finally, we present several applications in Discussion Sec.~\ref{Discussion}.

@@

In preparation for stating our main results, we first identify the combinatorial criteria on support sets of generating codes allowing for stable sparse 
representations.  Let $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$. 
% Let $\text{\rm Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_\ell\}$ be the $\mathbb{R}$-linear span of vectors $\mathbf{v}_1, \ldots, \mathbf{v}_\ell$. 
A vector $\mathbf{x}$ is said to be \emph{supported} on $S \subseteq [m]$ %, written supp$(\mathbf{x}) = S$, 
when $\mathbf{x} \in \text{\rm Span} \{\mathbf{e}_j\}_{j\in S}$, where $\mathbf{e}_j$ are the standard basis in $\mathbb R^m$.  Let $\mathbf{M}_j$ be the $j$th column of a matrix $\mathbf{M}$ and similarly let $\mathbf{M}_S$ be the submatrix formed by the columns of $\mathbf{M}$ indexed by $S$ (we set $\text{\rm Span}\{\mathbf{M}_\emptyset\} := \{\textbf{0}\}$).

@@

%The second ingredient necessary for the statement of Thm.~\ref{DeterministicUniquenessTheorem} 
Next, we describe a quantitative version of the spark condition.  % We first explain how the spark condition \eqref{SparkCondition} relates to t
The \emph{lower bound} \cite{Grcar10} of a matrix $\mathbf{A} \in \mathbb R^{n \times m}$ is the largest number $\alpha$ such that \mbox{$\|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$. By compactness of the unit sphere, every injective linear map has a nonzero lower bound; hence, if $\mathbf{A}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a nonzero lower bound. We therefore define the following domain-restricted lower bound:
\begin{align*}
L_k(\mathbf{A}) := \frac{1}{\sqrt{k}}\max \{ \alpha : \|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2 \text{ for all $k$-sparse } \mathbf{x}\}.
\end{align*}

@@

We also say that a set of $k$-sparse vectors is in \emph{general linear position} when any $k$ of them are linearly independent.
The following is the precise statement of our main result.

\begin{theorem}\label{DeterministicUniquenessTheorem}
%Fix integers $n, \ell$, and $k < m$. 
Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} for some $k < m$ and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ with the SIP.  Then there exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.

Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \hat m}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{\hat m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$ must have $\hat m \geq m$ and, provided $p = \hat m - \ell(\hat m - m)$ is positive, necessarily satisfy:
\begin{align}\label{Cstable}
\|(\mathbf{A}_J- \mathbf{B}_{\hat J} \mathbf{PD} )_j\|_2 \leq C_1 \varepsilon, \ \  \text{for all $j \in J$},
\end{align}
for some $J \in {[m] \choose p}$, $\hat J \in {[\hat m] \choose p}$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.

[ ** TODO ** write $\hat m$ version ] Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
\begin{align}\label{b-PDa}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for $i \in [N]$}.
\end{align}
\end{theorem}

@@

%Every matrix $\mathbf{B} \in \mathbb{R}^{n \times m}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
%\begin{align}\label{Cstable}
%\|\mathbf{A}_j - (\mathbf{B}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon,\ \ \   \text{for all $j \in [m]$},
%\end{align}
%for some permutation matrix $\mathbf{P}$ and invertible diagonal $\mathbf{D}$.

%Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
%\begin{align}\label{b-PDa}
%\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for all $i \in [N]$}.
%\end{align}
%\end{theorem}

%\begin{theorem}\label{DeterministicUniquenessTheorem}
%Fix integers $n, \ell$, and $k < m \leq \hat m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\hat m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.
%
%Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \hat m}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{\hat m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
%\begin{align}\label{Cstable}
%\|\mathbf{A}_{\phi(j)} - (\mathbf{B}_{J}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon,\ \ \   \text{for $j = 1, \ldots, |J|$},
%\end{align}
%%
%for some $J \subseteq [\hat m]$ of size \mbox{$\lfloor \hat m - \ell(\hat m - m) \rfloor$}, injective map $\phi$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
%%\footnote{We note that the condition $\varepsilon < L_2(\mathbf{A})$ above is necessary. When \mbox{$\mathbf{A}$ = $I$} and $\mathbf{x}_i = \mathbf{e}_i$, it turns out that $C_1 = 1$ and there is a $\mathbf{B}$ and $1$-sparse $\mathbf{\hat x}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$ violating \eqref{Cstable}.}
%
%Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}_J$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}_J\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
%\begin{align}\label{b-PDa}
%\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for all $i \in [N]$}.
%\end{align}
%\end{theorem}

%In other words, the smaller the difference $\hat m - m$, the more columns and coefficients of the original $n \times m$ dictionary $\mathbf{A}$ and $m$-dimensional codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \hat m$ dictionary $\mathbf{B}$ and $\hat m$-dimensional codes $\mathbf{\hat x}_i$. In particular, when $\hat m = m$ all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}.

>>>>

@@

\authordeclaration{The authors declare no conflict of interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: cjg@berkeley.edu, chillar@msri.org}

\keywords{Sparse linear coding, dictionary learning, uniqueness, compressed sensing}

@@

More formally, let $\mathbf{A} \in \mathbb R^{n \times m}$ be a matrix with columns $\mathbf{A}_j$ ($j = 1,\ldots,m$) and let dataset $Z$ consist of measurements:
\begin{align}\label{LinearModel}
\mathbf{z}_i = \mathbf{A}\mathbf{x}_i + \mathbf{n}_i,\ \ \  \text{$i=1,\ldots,N$},
\end{align}
for $k$-\emph{sparse} $\mathbf{x}_i \in \mathbb{R}^m$ having at most $k$ nonzero entries and \emph{noise} $\mathbf{n}_i \in \mathbb{R}^n$, with bounded norm $\| \mathbf{n}_i \|_2 \leq  \eta$ representing our combined worst-case uncertainty in  measuring $\mathbf{A}\mathbf{x}_i$.
The precise mathematical problem addressed here is the following.

\begin{problem}[Sparse linear coding]\label{InverseProblem}
Find a matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ such that $\|\mathbf{z}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \eta$ for all $i$.
\end{problem}

Note that any particular solution $(\mathbf{B}, \mathbf{\bar x}_1 \ldots, \mathbf{\bar x}_N)$ to this problem gives rise to an orbit of other solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal.  
We therefore consider whether solutions to Prob.~\ref{InverseProblem} are unique only up to this ambiguity.  More specifically, 
we seek conditions for when sparsely coded datasets have a stable unique representation.

\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse linear representation in $m$ dimensions} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. %Equivalently, $\mathbf{Y} = \mathbf{A}\mathbf{X}$ for $\mathbf{Y}$ and $\mathbf{X}$ having the $\mathbf{y}_i$ and $\mathbf{x}_i$ as their columns, respectively. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon,\ \ \   \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}

To see how Def. \ref{maindef} directly relates to Prob. \ref{InverseProblem}, suppose that $Y$
% \mbox{$Y = \{\mathbf{A} \mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$} 
has a stable $k$-sparse linear representation and fix $\delta_1, \delta_2$ to be the desired recovery accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then from the triangle inequality, it follows that any matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N$ solving Prob.~\ref{InverseProblem} are necessarily close to the original dictionary $\mathbf{A}$ and sparse codes $\mathbf{x}_i$.

@@

Our main technical finding is that a sufficient quantity of measurements generated from any dictionary satisfying the spark condition 
has a stable sparse representation.  More specifically, such dictionaries are uniquely identifiable from as few as \mbox{$N = m(k-1){m \choose k} + m$} sparse linear combinations of their columns up to an error that is linear in the measurement noise (Thm.~\ref{DeterministicUniquenessTheorem}). In fact, provided $(n, m, k)$ satisfy CS inequality \eqref{CScondition}, in almost all cases the dictionary learning problem is well-posed given enough data (Cor.~\ref{ProbabilisticCor}). Importantly, these guarantees hold without assuming the recovered matrix satisfies the spark condition. Our results  also apply when only an upper bound on the number of dictionary elements is known. The explicit, algorithm-independent criteria we provide should be a useful tool in the theory of sparse dictionary learning.

In the next section, we give precise statements of our main results, which include an explicit form for $\varepsilon(\delta_1, \delta_2)$. We then prove our main theorem (Thm.~\ref{DeterministicUniquenessTheorem}) in Sec.~\ref{DUT} after stating some additional definitions and lemmas required for the proof, including our main tool from combinatorial matrix analysis (Lem.~\ref{MainLemma}). All other proofs are relegated to the Supplemental. 
Finally, we present several applications in Discussion Sec.~\ref{Discussion}.

@@

Before stating main results, we first identify the combinatorial criteria on support sets of generating codes allowing for stable sparse 
representations.  Let $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$. 
% Let $\text{\rm Span}\{\mathbf{v}_1, \ldots, \mathbf{v}_\ell\}$ be the $\mathbb{R}$-linear span of vectors $\mathbf{v}_1, \ldots, \mathbf{v}_\ell$. 
A vector $\mathbf{x}$ is said to be \emph{supported} on $S \subseteq [m]$ %,
when $\mathbf{x} \in \text{\rm Span} \{\mathbf{e}_j\}_{j\in S}$, where $\mathbf{e}_j$ are the standard basis in $\mathbb R^m$.  We also write
 supp$(\mathbf{x}) = S$ when $S$ indexes the nonzero components of $\mathbf{x}$.
Let $\mathbf{M}_j$ be the $j$th column of a matrix $\mathbf{M}$ and similarly let $\mathbf{M}_S$ be the submatrix formed by the columns of $\mathbf{M}$ indexed by $S$ (we set $\text{\rm Span}\{\mathbf{M}_\emptyset\} := \{\textbf{0}\}$).

@@

%The second ingredient necessary for the statement of Thm.~\ref{DeterministicUniquenessTheorem} 
Next, we describe a quantitative version of the spark condition.  % We first explain how the spark condition \eqref{SparkCondition} relates to t
The \emph{lower bound} \cite{Grcar10} of a matrix $\mathbf{A} \in \mathbb R^{n \times m}$ is the largest number $\alpha$ such that \mbox{$\|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$. By compactness of the unit sphere, every injective linear map has a nonzero lower bound; hence, if $\mathbf{A}$ satisfies \eqref{SparkCondition}, then each submatrix formed from $2k$ of its columns or less has a nonzero lower bound. This motivates the following definition: % We therefore define the following domain-restricted lower bound:
\begin{align*}
L_k(\mathbf{A}) := \frac{1}{\sqrt{k}}\max \{ \alpha : \|\mathbf{A}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2 \text{ for all $k$-sparse } \mathbf{x}\}.
\end{align*}

@@

We say that a set of $k$-sparse vectors is in \emph{general linear position} when any $k$ of them are linearly independent.
The following is the precise statement of our main result.

\begin{theorem}\label{DeterministicUniquenessTheorem}
%Fix integers $n, \ell$, and $k < m$. 
Let $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfy \eqref{SparkCondition} for $k < m$, and suppose for some $\bar m$ more than \mbox{$(k-1){\bar m \choose k}$} $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} are supported in general linear position on each set in an $\ell$-regular $E \subseteq {[m] \choose k}$ with the SIP.  There is a constant $C_1 > 0$ for which the following hold for $\varepsilon < L_2(\mathbf{A}) / C_1$.

1.) Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{\bar m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$} for all $i$ must have $\bar m \geq m$ and, provided $p = \bar m - \ell(\bar m - m)$ is positive, necessarily satisfy:
\begin{align}\label{Cstable}
\|(\mathbf{A}_J- \mathbf{B}_{\bar J} \mathbf{PD} )_j\|_2 \leq C_1 \varepsilon, \ \  \text{for $j \in J$},
\end{align}
for some $J \in {[m] \choose p}$, $\bar J \in {[\bar m] \choose p}$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.

2.) Restricting dictionaries and sparse codes to respective $J, \bar{J}$, we furthermore have that if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
\begin{align}\label{b-PDa}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for $i \in [N]$}.
\end{align}
\end{theorem}

@@

%Every matrix $\mathbf{B} \in \mathbb{R}^{n \times m}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
%\begin{align}\label{Cstable}
%\|\mathbf{A}_j - (\mathbf{B}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon,\ \ \   \text{for all $j \in [m]$},
%\end{align}
%for some permutation matrix $\mathbf{P}$ and invertible diagonal $\mathbf{D}$.

%Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
%\begin{align}\label{b-PDa}
%\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for all $i \in [N]$}.
%\end{align}
%\end{theorem}

%\begin{theorem}\label{DeterministicUniquenessTheorem}
%Fix integers $n, \ell$, and $k < m \leq \bar m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies the spark condition \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\bar m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP.  There exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.
%
%Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \bar m}$ and $k$-sparse $\mathbf{\bar x}_1, \ldots, \mathbf{\bar x}_N \in \mathbb{R}^{\bar m}$ with  \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
%\begin{align}\label{Cstable}
%\|\mathbf{A}_{\phi(j)} - (\mathbf{B}_{J}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon,\ \ \   \text{for $j = 1, \ldots, |J|$},
%\end{align}
%%
%for some $J \subseteq [\bar m]$ of size \mbox{$\lfloor \bar m - \ell(\bar m - m) \rfloor$}, injective map $\phi$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
%%\footnote{We note that the condition $\varepsilon < L_2(\mathbf{A})$ above is necessary. When \mbox{$\mathbf{A}$ = $I$} and $\mathbf{x}_i = \mathbf{e}_i$, it turns out that $C_1 = 1$ and there is a $\mathbf{B}$ and $1$-sparse $\mathbf{\bar x}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\bar x}_i\|_2 \leq \varepsilon$ violating \eqref{Cstable}.}
%
%Furthermore, if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$, then $\mathbf{B}_J$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}_J\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$, and:
%\begin{align}\label{b-PDa}
%\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\bar x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,\ \ \   \text{for all $i \in [N]$}.
%\end{align}
%\end{theorem}

%In other words, the smaller the difference $\bar m - m$, the more columns and coefficients of the original $n \times m$ dictionary $\mathbf{A}$ and $m$-dimensional codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \bar m$ dictionary $\mathbf{B}$ and $\bar m$-dimensional codes $\mathbf{\bar x}_i$. In particular, when $\bar m = m$ all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}.

An important consequence of Thm.~\ref{DeterministicUniquenessTheorem} is that \eqref{def1} is guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: % [*** TODO: verify this is still correct ***]
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \max_{i \in [N]} \|\mathbf{x}_i\|_1  + \delta_2 \right) } \right\}.
\end{align}
For expositional clarity, we delay defining the (explicit) constant $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E)$ until Section \ref{DUT} (\eqref{Cdef1}).  % , since it requires additional definitions not critical to the statement of our results.
